% !TeX root = ThesisMain.tex
% !TeX program = XeLaTeX
% !TeX encoding = UTF-8
% !TeX spellcheck = en_GB

\documentclass[../ThesisMain]{subfiles}
\ifSubfilesClassLoaded{}{}%

\begin{document}
\doublespacing%
\begin{acknowledgements}
I would like to express my deepest gratitude to my supervisor, Professor Budimir Rosic, for his invaluable guidance, continuous support, and patience throughout this research. His expertise in turbomachinery and computational fluid dynamics has been instrumental in shaping this work, and his encouragement has been a constant source of motivation.

I am grateful to St Anne's College for providing a welcoming home during my time at Oxford. The college community has been a source of friendship, support, and countless memorable moments that made this journey so much richer.

I would like to thank the Oxford Thermofluids Institute for providing an exceptional research environment. The technical staff deserve special recognition for maintaining the high-performance computing facilities essential for the extensive simulations and machine learning experiments in this thesis. Their expertise and responsiveness made this computationally demanding work possible.

I am also thankful for the friendships formed across the collegiate university---from formal halls to late-night discussions, my friends from St Anne's and other colleges have provided welcome respite from research and reminded me that there is life beyond the lab.

Finally, I extend my heartfelt thanks to my family for their unwavering love and support throughout my doctoral studies. Their belief in me, even when I doubted myself, has been the foundation upon which this work was built.

\end{acknowledgements}
\newpage
\end{document}
